---
title: "EIA_MER_DataAccess"
author: "Paul Leiby"
date: "Aug 09, 2018"
output: word_document
---

File: EIA_MER_DataAccess_2018.Rmd

Created by: Paul N. Leiby
Last revised: 20190809

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Set working environment and paths

```{r setWorkingEnv}
## clear: remove (almost) everything in the working environment.
## You will get no warning, so don't do this unless you are really sure.
rm(list = ls())

#worksubpath = "/SPR/Data/EIA_MER"
datasubpath = "../"
```

This reads from the EIA data browser website for the Annual Energy Review and Monthly Energy Review

MER Data Section List
-----------------

1. Section 1. Energy Overview
2. Section 2. Energy Consumption by Sector
3. Section 3. Petroleum
4. Section 4. Natural Gas
5. Section 5. Crude Oil and Natural Gas Resource Development
6. Section 6. Coal
7. Section 7. Electricity
8. Section 8. Nuclear Energy
9. Section 9. Energy Prices
10. Section 10. Renewable Energy
11. Section 11. International Petroleum
12. Section 12. Environment
13. Appendix A. British Thermal Unit Conversion Factors
14. Appendix B. Metric Conversion Factors, Metric Prefixes, and Other Physical Conversion Factors
15. Appendix C. Population, U.S. Gross Domestic Product, and U.S. Gross Output


MER Data Table List
--------------

    Section 1. Energy Overview
    1.2 Primary Energy Production by Source
    1.1 Primary Energy Overview
    1.3 Primary Energy Consumption by Source
    1.4a Primary Energy Imports by Source
    1.4b Primary Energy Exports by Source and Total Net Imports
    1.5 Merchandise Trade Value
    1.6 Cost of Fuels to End Users in Real (1982-1984) Dollars
    1.7 Primary Energy Consumption, Energy Expenditures, and Carbon Dioxide Emissions Indicators
    1.8 Motor Vehicle Mileage, Fuel Consumption, and Fuel Economy
    1.9 Heating Degree-Days by Census Division
    1.10 Cooling Degree-Days by Census Division
    Section 2. Energy Consumption by Sector
    2.1 Energy Consumption by Sector
    2.2 Residential Sector Energy Consumption
    2.3 Commercial Sector Energy Consumption
    2.4 Industrial Sector Energy Consumption
    2.5 Transportation Sector Energy Consumption
    2.6 Electric Power Sector Energy Consumption
    2.7 U.S. Government Energy Consumption by Agency, Fiscal Years
    2.8 U.S. Government Energy Consumption by Source, Fiscal Years
    Section 3. Petroleum
    3.1 Petroleum Overview
    3.2 Refinery and Blender Net Inputs and Net Production
    3.3 Petroleum Trade
    3.3a Overview
    3.3b Imports and Exports by Type
    3.3c Imports From OPEC Countries
    3.3d Imports From Non-OPEC Countries
    3.4 Petroleum Stocks
    3.5 Petroleum Products Supplied by Type
    3.6 Heat Content of Petroleum Products Supplied by Type
    3.7 Petroleum Consumption
    3.7a Residential and Commercial Sectors
    3.7b Industrial Sector
    3.7c Transportation and Electric Power Sectors
    3.8 Heat Content of Petroleum Consumption
    3.8a Residential and Commercial Sectors
    3.8b Industrial Sector
    3.8c Transportation and Electric Power Sectors
    Section 4. Natural Gas
    4.1 Natural Gas Overview
    4.2 Natural Gas Trade by Countr
    4.3 Natural Gas Consumption by Sector
    4.4 Natural Gas in Underground Storage
    Section 5. Crude Oil and Natural Gas Resource Development
    5.1 Crude Oil and Natural Gas Drilling Activity Measurements
    5.2 Crude Oil and Natural Gas Exploratory and Development Wells
    Section 6. Coal
    6.1 Coal Overview
    6.2 Coal Consumption by Sector
    6.3 Coal Stocks by Sector
    Section 7. Electricity
    7.1 Electricity Overview
    7.2 Electricity Net Generation
    7.2a Total (All Sectors)
    7.2b Electric Power Sector
    7.2c Commercial and Industrial Sectors
    7.3 Consumption of Combustible Fuels for Electricity Generation
    7.3a Total (All Sectors)
    7.3b Electric Power Sector
    7.3c Commercial and Industrial Sectors (Selected Fuels)
    7.4 Consumption of Combustible Fuels for Electricity Generation and Useful Thermal Output
    7.4a Total (All Sectors)
    7.4b Electric Power Sector
    7.4c Commercial and Industrial Sectors (Selected Fuels)
    7.5 Stocks of Coal and Petroleum: Electric Power Sector
    7.6 Electricity End Use
    Section 8. Nuclear Energy
    8.1 Nuclear Energy Overview
    Section 9. Energy Prices
    9.1 Crude Oil Price Summary
    9.2 F.O.B. Costs of Crude Oil Imports From Selected Countries
    9.3 Landed Costs of Crude Oil Imports From Selected Countries
    9.4 Retail Motor Gasoline and On-Highway Diesel Fuel Prices
    9.5 Refiner Prices of Residual Fuel Oil
    9.6 Refiner Prices of Petroleum Products for Resale
    9.7 Refiner Prices of Petroleum Products to End Users
    9.8 Average Retail Prices of Electricity
    9.9 Cost of Fossil-Fuel Receipts at Electric Generating Plants
    9.10 Natural Gas Prices
    Section 10. Renewable Energy
    10.1 Renewable Energy Production and Consumption by Source
    10.2 Renewable Energy Consumption
    10.2a Residential and Commercial Sectors
    10.2b Industrial and Transportation Sectors
    10.2c Electric Power Secto
    10.3 Fuel Ethanol Overview
    10.4 Biodiesel and Other Renewable Fuels Overview
    Section 11. International Petroleum
    11.1 World Crude Oil Production
    11.1a OPEC Members
    11.1b Persian Gulf Nations, Non-OPEC, and World
    11.2 Petroleum Consumption in OECD Countries
    11.3 Petroleum Stocks in OECD Countries
    Section 12. Environment
    12.1 Carbon Dioxide Emissions From Energy Consumption by Sourc
    12.2 Carbon Dioxide Emissions From Energy Consumption: Residential Secto
    12.3 Carbon Dioxide Emissions From Energy Consumption: Commercial Sector
    12.4 Carbon Dioxide Emissions From Energy Consumption: Industrial Sector
    12.5 Carbon Dioxide Emissions From Energy Consumption: Transportation Sector
    12.6 Carbon Dioxide Emissions From Energy Consumption: Electric Power Sector
    12.7 Carbon Dioxide Emissions From Biomass Energy Consumptio
    Appendix A. British Thermal Unit Conversion Factors
    A1. Approximate Heat Content of Petroleum and Other Liquids
    A2. Approximate Heat Content of Petroleum Production, Imports, and Exports
    A3. Approximate Heat Content of Petroleum Consumption and Fuel Ethanol
    A4. Approximate Heat Content of Natural Gas
    A5. Approximate Heat Content of Coal and Coal Coke
    A6. Approximate Heat Rates for Electricity, and Heat Content of Electricity
    Appendix B. Metric Conversion Factors, Metric Prefixes, and Other Physical Conversion Factors
    B1. Metric Conversion Factors
    B2. Metric Prefixes
    B3. Other Physical Conversion Factors
    Appendix C. Population, U.S. Gross Domestic Product, and U.S. Gross Output
    C1. Population, U.S. Gross Domestic Product, and U.S. Gross Output
    Appendix D. Estimated Primary Energy Consumption in the United States, Selected Years, 1635-1945
    D1. Estimated Primary Energy Consumption in the United States, Selected Years, 1635-1945

Examples of Reading Tables from MER Data Site
------------------------

* Requires internet connection or local CSV file availability

```{r setupWorkspace}
#library(XLConnect)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)

```

```{r readMERdata}
# set working directory - not necessary for RStudio project approach, or if console session launched from working directory
# setwd(paste0(workroot,worksubpath))

# Specify data location
# EIA MER URL query, less table name
# MER_URLbase =  "https://www.eia.gov/totalenergy/data/browser/csv.cfm?tbl=" # outdated
MER_URLbase =  "https://www.eia.gov/totalenergy/data/browser/csv.php?tbl="
# e.g.: sourceURL =  "http://www.eia.gov/totalenergy/data/browser/csv.cfm?tbl=T11.01A"

# download latest version of 4 tables from EIA/MER website
naStrings = c("","Not Available","Not Applicable") # ways that MER indicates NA

#tableSiteURL = "https://www.eia.gov/totalenergy/data/browser/"
tableSiteURL = "https://www.eia.gov/totalenergy/data/browser/"
#csvExtension = "csv.cfm?tbl="
#xlsExtension = "xls.cfm?tbl="
csvExtension = "csv.php?tbl="
xlsExtension = "xls.php?tbl="


# "http://www.eia.gov/totalenergy/data/browser/csv.cfm?tbl=T01.03" # MER Table 1.3 Primary Energy Consumption by Source
# "http://www.eia.gov/totalenergy/data/browser/csv.cfm?tbl=T01.06"
# "http://www.eia.gov/totalenergy/data/browser/xls.cfm?tbl=T01.06"
# "http://www.eia.gov/totalenergy/data/browser/csv.cfm?tbl=T01.08"
# "http://www.eia.gov/totalenergy/data/browser/csv.cfm?tbl=T03.07C" # annual and monthly

# syntax for call to EIA data browser
browserSiteURL = "https://www.eia.gov/totalenergy/data/browser/index.php?tbl=T01.03#/?f=A"
# http://www.eia.gov/beta/MER/index.cfm?tbl=T03.07C#/?f=M&start=197301&end=201603&charted=8-13

onlineReading = T

# sectionWanted = "T03"
# tableWanted = "07C"
sectionWanted = "T01"
tableWanted = "03"

tableURL = paste0(tableSiteURL, csvExtension, sectionWanted, ".", tableWanted)

readMERtable <- function(sectionW, tableW, onlineReading = T) {
  if(onlineReading==T) { # get info from URL for EIA MER data site
  #  T01.03 = read.csv(url(tableURL))
    newdf = read.csv(url(paste0(tableSiteURL,csvExtension, sectionW, ".", tableW)))
  } else { # assume csv file is locally available
  #  T01.03 = read.csv(table))
    newdf = read.csv(    paste0(workroot,datasubpath, "MER_", sectionW, "_", tableW,".csv"))
  }
    
  newdf$Value = as.numeric(as.character(newdf$Value))
  return(newdf)
}

# First drop the values and dates, to get a table of metadata:
#  variable names (MSN), and associated descriptions and units
getMetaData <- function(MERdf, tableName) {
  md = MERdf %>% select(-c(YYYYMM, Value)) %>% unique()
  # add table name, so that we can combine datasets and reconstruct tables
  md$Table = tableName
  row.names(md) <- c()
  return(md)
}

```

```{r preprocessDataT01_03}
onlineReading = T
sectionWanted = "T01"
tableWanted = "03"

EIAdf = readMERtable(sectionWanted, tableWanted, onlineReading)

newTableName = paste("MER",sectionWanted, tableWanted,sep="_")
assign(newTableName, EIAdf)

EIAdf_metadata = data.frame()
EIAdf_metadata = rbind(EIAdf_metadata, getMetaData(EIAdf, newTableName ))
```

```{r}
# for table formatting, see 
# https://cran.r-project.org/web/packages/gridExtra/vignettes/tableGrob.html
# https://github.com/baptiste/gridextra/wiki/tableGrob
library(grid)
library(gridExtra)
# rescale parts of the table
mytheme <- gridExtra::ttheme_default(
    core =    list(fg_params=list(cex = 0.5)),
    colhead = list(fg_params=list(cex = 1.0)),
    rowhead = list(fg_params=list(cex = 1.0)))
myt <- gridExtra::tableGrob(EIAdf_metadata, theme = mytheme)
grid.draw(myt)
# grid.table(EIAdf_metadata, gp=gpar(fontsize=8)) # throws error for fontsize param

# produce wide table with one column per variable, sorted by date
EIAdf_w = EIAdf %>% select(-c(Column_Order, Description, Unit)) %>% 
  spread(key = MSN, value=Value) %>% arrange(YYYYMM)
summary(EIAdf_w)
```

Visualize the data
-------------------

```{r visualizeData0}

```

```{r defineDownloadEIA_MER}

downloadEIA_MER_table <- function (sectionW, tableW) {
  # function to download specified table from EIA's MER,
  #  create a dataframe named appropriately,
  #  and add table metadata to a dataframe compiling that info
  # Use suppressWarnings() to avoid "NAs introduced by coercion" message
  EIAdf = readMERtable(sectionW, tableW, onlineReading=T)
  tableName = paste("MER",sectionW, tableW,sep="_")
  assign(tableName, EIAdf)

  # to check if dataframe exists, given name as text string df_name:
  #  exists(df_name) && is.data.frame(get(df_name))
  if (!exists("EIAdf_metadata")) {
      EIAdf_metadata = data.frame()  
  }

  EIAdf_metadata = rbind(df_metadata, getMetaData(EIAdf, tableName ))
  
}

```

### Get US Oil Production, Consumption, and Import Data

```{r preprocessDataT03_01}
onlineReading = T

sectionWanted = "T03"
tableWanted = "01"
# Use suppressWarnings() to avoid "NAs introduced by coercion" message
EIAdf = readMERtable(sectionWanted, tableWanted, onlineReading)
newTableName = paste("MER",sectionWanted, tableWanted,sep="_")
assign(newTableName, EIAdf)
# to check if dataframe exists, given name as text string df_name:
#  exists(df_name) && is.data.frame(get(df_name))
if (!exists("EIAdf_metadata")) {
  EIAdf_metadata = data.frame()  
}
EIAdf_metadata = rbind(EIAdf_metadata, getMetaData(EIAdf, newTableName ))


sectionWanted = "T11"
tableWanted = "02"
# Use suppressWarnings() to avoid "NAs introduced by coercion" message
EIAdf = readMERtable(sectionWanted, tableWanted, onlineReading)
newTableName = paste("MER",sectionWanted, tableWanted,sep="_")
assign(newTableName, EIAdf)
# to check if dataframe exists, given name as text string df_name:
#  exists(df_name) && is.data.frame(get(df_name))
if (!exists("EIAdf_metadata")) {
  EIAdf_metadata = data.frame()  
}
EIAdf_metadata = rbind(EIAdf_metadata, getMetaData(EIAdf, newTableName ))


sectionWanted = "T11"
tableWanted = "01B"
# Use suppressWarnings() to avoid "NAs introduced by coercion" message
EIAdf = readMERtable(sectionWanted, tableWanted, onlineReading)
newTableName = paste("MER",sectionWanted, tableWanted,sep="_")
assign(newTableName, EIAdf)
# to check if dataframe exists, given name as text string df_name:
#  exists(df_name) && is.data.frame(get(df_name))
if (!exists("EIAdf_metadata")) {
  EIAdf_metadata = data.frame()  
}
EIAdf_metadata = rbind(EIAdf_metadata, getMetaData(EIAdf, newTableName ))

sectionWanted = "T11"
tableWanted = "01A"
# Use suppressWarnings() to avoid "NAs introduced by coercion" message
EIAdf = readMERtable(sectionWanted, tableWanted, onlineReading)
newTableName = paste("MER",sectionWanted, tableWanted,sep="_")
assign(newTableName, EIAdf)
# to check if dataframe exists, given name as text string df_name:
#  exists(df_name) && is.data.frame(get(df_name))
if (!exists("EIAdf_metadata")) {
  EIAdf_metadata = data.frame()  
}
EIAdf_metadata = rbind(EIAdf_metadata, getMetaData(EIAdf, newTableName ))

```

### Combine the read Tables 

```{r combineReadTables}

# combine the petroleum data table
MER_PetData = rbind(MER_T03_01, MER_T11_01A, MER_T11_01B, MER_T11_02)
MER_PetData$YYYYMMc = as.character(MER_PetData$YYYYMM)

MER_PetData = MER_PetData %>% 
  select(-Column_Order) %>% 
  separate(YYYYMMc, sep= 4, into = c("YYYY", "MM")) # split into year and month fields

```

```{r}
# look for duplicate variables and drop (necessary before spread(), to avoid error)

# Remove duplicate rows based on one or more col values: dplyr::distinct(my_data, colname).
#   distinct() is best-suited for interactive use. 
# The function distinct_() should be used for calling from a function.
# In this case the input must be “quoted”. 
# Also can use base fns unique() and duplicated()
# MER_PetData = MER_PetData %>%
#  distinct_("MSN", "Value", "YYYYMM")  # leaves only the variables tested for distinctness

Pet = MER_PetData %>% select(MSN, Value, YYYYMM)
MER_PetData = MER_PetData[!duplicated(Pet), ]  # keep all vars, if distinct name/value/date
```

get annual data subset

```{r}
MER_PetDataA = MER_PetData %>% filter(MM == "13") # get annual data only

# MER_PetDataA = MER_PetDataA %>%
#  distinct_("MSN", "Value", "YYYYMM")

```

Keep only essential fields (name MSN, Value, and date )
 and spread out vars into columns

```{r}

varsWanted = c("PAPRPUS", "NLPRPUS", "RYPRPUS", 
               "PAGLPUS", "PASCPUS", "PNPRPUS", 
               "PAEXPUS", "PAIMPUS", "PANIPUS", "PATCPUS")
EIAdf_metadata %>% filter(MSN %in% varsWanted)
# PATCPUS = Petroleum Consumption, United States = Petroleum Products Supplied
# PAPRPUS = Crude Oil Production, United States = Crude Oil Production, Total
# Petroleum Consumption Series From Tables 3.7A, B, C
# c("PARCPUS", "PACCPUS", "PAICPUS","PAACPUS", "PAEIZUS")
# Residential, Commercial, Industrial, Transportation and Electric

PetA = MER_PetDataA %>% 
  select(-c(YYYYMM, Description, Unit, MM)) %>%
  filter(MSN %in% varsWanted) %>%
  spread(key = MSN, value=Value)
```

```{r}
rownames(PetA) = PetA$YYYY
cat("2007 Consumption: ", PetA["2007", "PATCPUS"])
cat("2012 Consumption: ", PetA["2012", "PATCPUS"])
cat("2017 Consumption: ", PetA["2017", "PATCPUS"])

cat("2007 Production: ", PetA["2007", "PATCPUS"])
cat("2012 Production: ", PetA["2012", "PATCPUS"])
cat("2017 Production: ", PetA["2017", "PATCPUS"])
```


### Plot Monthly Petroleum Consumption

``` {r plotPetrStocksumption}
PetrCons = T11_02
PetrCons$Date <- as.Date(paste0(as.character(PetrCons$YYYYMM),"15"),format="%Y%m%d")
PetrCons = PetrCons %>% separate(Description, sep=",", c("Series", "Country"))

PetrCons %>% filter(Series=="Petroleum Consumption", !is.na(Value)) %>%
  ggplot(aes(x=Date, y=Value, color=Country)) + 
  geom_line() + ggtitle("Petroleum Consumption") + ylab("1000 B/day")

```

### Prices

    9.1 Crude Oil Price Summary
    9.2 F.O.B. Costs of Crude Oil Imports From Selected Countries
    9.3 Landed Costs of Crude Oil Imports From Selected Countries
    9.4 Retail Motor Gasoline and On-Highway Diesel Fuel Prices
    9.6 Refiner Prices of Petroleum Products for Resale
    9.7 Refiner Prices of Petroleum Products to End Users
    9.10 Natural Gas Prices
    10.1 Renewable Energy Production and Consumption by Source
    10.2 Renewable Energy Consumption
    10.3 Fuel Ethanol Overview

```{r readMERPrices}
sectionWanted = "T09" # Petroleum Prices
tableWanted = "01"    # Crude Oil Price Summary
EIAdf = readMERtable(sectionWanted, tableWanted, onlineReading)

assign(paste("MER",sectionWanted, tableWanted,sep="_"), EIAdf)

```


### Plot Monthly OECD Stocks

``` {r plotPetrStocks}
PetrStocks = T11_03
PetrStocks$Date <- as.Date(paste0(as.character(PetrStocks$YYYYMM),"15"),format="%Y%m%d")
PetrStocks = PetrStocks %>% separate(Description, sep=",", c("Series", "Country"))

PetrStocks %>% filter(Series=="Petroleum Stocks", !is.na(Value)) %>%
  ggplot(aes(x=Date, y=Value, color=Country)) + 
  geom_line() + ggtitle("Petroleum Stocks") + ylab("Million Barrels")

```
