---
title: "EIA_MER_DataAccess"
author: "Paul Leiby"
output: word_document
---

File: EIA_MER_DataAccess.Rmd

Created by: Paul N. Leiby
Last revised: 20220302

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Set working environment and paths

```{r setWorkingEnv, echo=F}
## clear: remove (almost) everything in the working environment.
## You will get no warning, so don't do this unless you are really sure.
rm(list = ls())

echoWorking = T    # flag to control display of code execution
#worksubpath = "/SPR/Data/EIA_MER"
datasubpath = "../data/" # local relative reference to project directory
```

This reads from the EIA data browser website for the Annual Energy Review and Monthly Energy Review

MER Data Section List
-----------------

1. Section 1. Energy Overview
2. Section 2. Energy Consumption by Sector
3. Section 3. Petroleum
4. Section 4. Natural Gas
5. Section 5. Crude Oil and Natural Gas Resource Development
6. Section 6. Coal
7. Section 7. Electricity
8. Section 8. Nuclear Energy
9. Section 9. Energy Prices
10. Section 10. Renewable Energy
11. Section 11. International Petroleum
12. Section 12. Environment
13. Appendix A. British Thermal Unit Conversion Factors
14. Appendix B. Metric Conversion Factors, Metric Prefixes, and Other Physical Conversion Factors
15. Appendix C. Population, U.S. Gross Domestic Product, and U.S. Gross Output


MER Data Table List
--------------

    Section 1. Energy Overview
    1.2 Primary Energy Production by Source
    1.1 Primary Energy Overview
    1.3 Primary Energy Consumption by Source
    1.4a Primary Energy Imports by Source
    1.4b Primary Energy Exports by Source and Total Net Imports
    1.5 Merchandise Trade Value
    1.6 Cost of Fuels to End Users in Real (1982-1984) Dollars
    1.7 Primary Energy Consumption, Energy Expenditures, and Carbon Dioxide Emissions Indicators
    1.8 Motor Vehicle Mileage, Fuel Consumption, and Fuel Economy
    1.9 Heating Degree-Days by Census Division
    1.10 Cooling Degree-Days by Census Division
    Section 2. Energy Consumption by Sector
    2.1 Energy Consumption by Sector
    2.2 Residential Sector Energy Consumption
    2.3 Commercial Sector Energy Consumption
    2.4 Industrial Sector Energy Consumption
    2.5 Transportation Sector Energy Consumption
    2.6 Electric Power Sector Energy Consumption
    2.7 U.S. Government Energy Consumption by Agency, Fiscal Years
    2.8 U.S. Government Energy Consumption by Source, Fiscal Years
    Section 3. Petroleum
    3.1 Petroleum Overview
    3.2 Refinery and Blender Net Inputs and Net Production
    3.3 Petroleum Trade
    3.3a Overview
    3.3b Imports and Exports by Type
    3.3c Imports From OPEC Countries
    3.3d Imports From Non-OPEC Countries
    3.4 Petroleum Stocks
    3.5 Petroleum Products Supplied by Type
    3.6 Heat Content of Petroleum Products Supplied by Type
    3.7 Petroleum Consumption
    3.7a Residential and Commercial Sectors
    3.7b Industrial Sector
    3.7c Transportation and Electric Power Sectors
    3.8 Heat Content of Petroleum Consumption
    3.8a Residential and Commercial Sectors
    3.8b Industrial Sector
    3.8c Transportation and Electric Power Sectors
    Section 4. Natural Gas
    4.1 Natural Gas Overview
    4.2 Natural Gas Trade by Countr
    4.3 Natural Gas Consumption by Sector
    4.4 Natural Gas in Underground Storage
    Section 5. Crude Oil and Natural Gas Resource Development
    5.1 Crude Oil and Natural Gas Drilling Activity Measurements
    5.2 Crude Oil and Natural Gas Exploratory and Development Wells
    Section 6. Coal
    6.1 Coal Overview
    6.2 Coal Consumption by Sector
    6.3 Coal Stocks by Sector
    Section 7. Electricity
    7.1 Electricity Overview
    7.2 Electricity Net Generation
    7.2a Total (All Sectors)
    7.2b Electric Power Sector
    7.2c Commercial and Industrial Sectors
    7.3 Consumption of Combustible Fuels for Electricity Generation
    7.3a Total (All Sectors)
    7.3b Electric Power Sector
    7.3c Commercial and Industrial Sectors (Selected Fuels)
    7.4 Consumption of Combustible Fuels for Electricity Generation and Useful Thermal Output
    7.4a Total (All Sectors)
    7.4b Electric Power Sector
    7.4c Commercial and Industrial Sectors (Selected Fuels)
    7.5 Stocks of Coal and Petroleum: Electric Power Sector
    7.6 Electricity End Use
    Section 8. Nuclear Energy
    8.1 Nuclear Energy Overview
    Section 9. Energy Prices
    9.1 Crude Oil Price Summary
    9.2 F.O.B. Costs of Crude Oil Imports From Selected Countries
    9.3 Landed Costs of Crude Oil Imports From Selected Countries
    9.4 Retail Motor Gasoline and On-Highway Diesel Fuel Prices
    9.5 Refiner Prices of Residual Fuel Oil
    9.6 Refiner Prices of Petroleum Products for Resale
    9.7 Refiner Prices of Petroleum Products to End Users
    9.8 Average Retail Prices of Electricity
    9.9 Cost of Fossil-Fuel Receipts at Electric Generating Plants
    9.10 Natural Gas Prices
    Section 10. Renewable Energy
    10.1 Renewable Energy Production and Consumption by Source
    10.2 Renewable Energy Consumption
    10.2a Residential and Commercial Sectors
    10.2b Industrial and Transportation Sectors
    10.2c Electric Power Secto
    10.3 Fuel Ethanol Overview
    10.4 Biodiesel and Other Renewable Fuels Overview
    Section 11. International Petroleum
    11.1 World Crude Oil Production
    11.1a OPEC Members
    11.1b Persian Gulf Nations, Non-OPEC, and World
    11.2 Petroleum Consumption in OECD Countries
    11.3 Petroleum Stocks in OECD Countries
    Section 12. Environment
    12.1 Carbon Dioxide Emissions From Energy Consumption by Sourc
    12.2 Carbon Dioxide Emissions From Energy Consumption: Residential Secto
    12.3 Carbon Dioxide Emissions From Energy Consumption: Commercial Sector
    12.4 Carbon Dioxide Emissions From Energy Consumption: Industrial Sector
    12.5 Carbon Dioxide Emissions From Energy Consumption: Transportation Sector
    12.6 Carbon Dioxide Emissions From Energy Consumption: Electric Power Sector
    12.7 Carbon Dioxide Emissions From Biomass Energy Consumptio
    Appendix A. British Thermal Unit Conversion Factors
    A1. Approximate Heat Content of Petroleum and Other Liquids
    A2. Approximate Heat Content of Petroleum Production, Imports, and Exports
    A3. Approximate Heat Content of Petroleum Consumption and Fuel Ethanol
    A4. Approximate Heat Content of Natural Gas
    A5. Approximate Heat Content of Coal and Coal Coke
    A6. Approximate Heat Rates for Electricity, and Heat Content of Electricity
    Appendix B. Metric Conversion Factors, Metric Prefixes, and Other Physical Conversion Factors
    B1. Metric Conversion Factors
    B2. Metric Prefixes
    B3. Other Physical Conversion Factors
    Appendix C. Population, U.S. Gross Domestic Product, and U.S. Gross Output
    C1. Population, U.S. Gross Domestic Product, and U.S. Gross Output
    Appendix D. Estimated Primary Energy Consumption in the United States, Selected Years, 1635-1945
    D1. Estimated Primary Energy Consumption in the United States, Selected Years, 1635-1945

Examples of Reading Tables from MER Data Site
------------------------

* Requires internet connection or local CSV file availability

```{r setupWorkspace, echo=echoWorking}
#library(XLConnect)
library(tidyverse)
# library(dplyr)
# library(lubridate)
library(knitr)   # for knitr::kable
library(grid)
library(gridExtra) # for gridExtra::tableGrob formatted table display
```

### Establish Locations (URLs) and Functions for Reading MER Data)

```{r readMERdata_setup, echo=echoWorking}
# set working directory - not necessary for RStudio project approach,
# or if console session launched from working directory
# setwd(paste0(workroot,worksubpath))

# Specify data location
# EIA MER URL query, less table name
# MER_URLbase =  "https://www.eia.gov/totalenergy/data/browser/csv.cfm?tbl=" # outdated
MER_URLbase =  "https://www.eia.gov/totalenergy/data/browser/csv.php?tbl="
# e.g.: sourceURL =  "https://www.eia.gov/totalenergy/data/browser/csv.php?tbl=T11.01A"

naStrings = c("","Not Available","Not Applicable") # ways that MER indicates NA

#tableSiteURL = "https://www.eia.gov/totalenergy/data/browser/"
tableSiteURL = "https://www.eia.gov/totalenergy/data/browser/"

#csvExtension = "csv.cfm?tbl="
#xlsExtension = "xls.cfm?tbl="
csvExtension = "csv.php?tbl="
xlsExtension = "xls.php?tbl="


# download latest version of 4 tables from EIA/MER website
# "http://www.eia.gov/totalenergy/data/browser/csv.cfm?tbl=T01.03" # MER Table 1.3 Primary Energy Consumption by Source
# "http://www.eia.gov/totalenergy/data/browser/csv.cfm?tbl=T01.06"
# "http://www.eia.gov/totalenergy/data/browser/xls.cfm?tbl=T01.06"
# "http://www.eia.gov/totalenergy/data/browser/csv.cfm?tbl=T01.08"
# "http://www.eia.gov/totalenergy/data/browser/csv.cfm?tbl=T03.07C" # annual and monthly

# syntax for call to EIA data browser
browserSiteURL = "https://www.eia.gov/totalenergy/data/browser/index.php?tbl=T01.03#/?f=A"

onlineReading = T  # flag indicating whether to download from EIA site (vs read local copy of file)

# tableURL = paste0(tableSiteURL, csvExtension, sectionWanted, ".", tableWanted)
```


```{r defFn_readMERdata, echo=echoWorking}
# Read all MER tables as csv files,
#   either downloading from EIA site or locally, based on onlineReading flag
readMERtable <- function(sectionW, tableW, onlineReading = T) {
  # function to download specified table from EIA's MER,
  #  create a dataframe named appropriately,
  #  and add table metadata to a dataframe compiling that info
  # Use suppressWarnings() to avoid "NAs introduced by coercion" message
  if(onlineReading==T) { # get info from URL for EIA MER data site
    if (tableW !="") {
      #  as for table T01.03 
      newdf = read.csv(url(paste0(tableSiteURL,csvExtension, sectionW, ".", tableW)))
    } else { # special treatment of appendix tables the do not use "." separator or tableW field
      #  as for table "TC1"
      newdf = read.csv(url(paste0(tableSiteURL,csvExtension, sectionW)))
    }
  } else { # assume csv file is locally available
  #  e.g. T01.03 = read.csv("MER_T01_03.csv"))
    newdf = read.csv(    paste0(workroot,datasubpath, "MER_", sectionW, "_", tableW,".csv"))
  }
    
  newdf$Value = as.numeric(as.character(newdf$Value))
  return(newdf)
}
```


```{r defFn_getMetadata, echo=echoWorking}
getMetaData <- function(MERdf, tableName) {
  # fn to Extract a table of metadata from an EIA MER table dataframe
  #  Drop the values and dates, leaving
  #  variable names (MSN), and associated descriptions and units
  md = MERdf %>% select(-c(YYYYMM, Value)) %>% unique()
  # add table name, so that we can combine datasets and reconstruct tables
  md$Table = tableName
  row.names(md) <- c()
  return(md)
}

```

The EIA table browser for the _Monthly Energy Report_ (MER) is found at `r tableSiteURL`.

```{r specifyTableT01_03, echo=echoWorking}
onlineReading = T
sectionWanted = "T03"
tableWanted = "01"
```

Example: Use defined functions `readMERdata` and `getMetaData` to
load a single table, `r paste("MER",sectionWanted, tableWanted,sep="_")`, from the online MER:
Also get Metadata for this table

```{r preprocessDataT01_03, echo=echoWorking}
EIAdf = readMERtable(sectionWanted, tableWanted, onlineReading)

## Warning in readMERtable(sectionWanted, tableWanted, onlineReading) :
##   NAs introduced by coercion
  
newTableName = paste("MER",sectionWanted, tableWanted,sep="_")
assign(newTableName, EIAdf) # rename default table name from EIAdf to actual

EIAdf_metadata = data.frame()
EIAdf_metadata = rbind(EIAdf_metadata, getMetaData(EIAdf, newTableName ))
```

```{r showEIAdf_table, echo=echoWorking}
# for table formatting, see 
# https://cran.r-project.org/web/packages/gridExtra/vignettes/tableGrob.html
# https://github.com/baptiste/gridextra/wiki/tableGrob
# rescale parts of the table
mytheme <- gridExtra::ttheme_default(
    core =    list(fg_params=list(cex = 0.5)),
    colhead = list(fg_params=list(cex = 1.0)),
    rowhead = list(fg_params=list(cex = 1.0)))
myt <- gridExtra::tableGrob(select(EIAdf_metadata,-Column_Order), theme = mytheme)
grid.draw(myt)
# grid.table(EIAdf_metadata, gp=gpar(fontsize=8)) # throws error for fontsize param

# produce wide table with one column per variable, sorted by date
EIAdf_w = EIAdf %>% select(-c(Column_Order, Description, Unit)) %>% 
  spread(key = MSN, value=Value) %>% arrange(YYYYMM)
summary(EIAdf_w)
```

Create Functions for Generalized Download of MER Data
-------------------

Data are coded with dates (month 13 denotes annual value), and metadata indicating the series description and units.
Date are in long (in dplyr terms, "gather"ed) form.


```{r defFn_downloadEIA_MER, echo=echoWorking}
readMERtable_and_rename <- function (sectionW, tableW) {
  # function to download specified table from EIA's MER,
  #  create a dataframe named appropriately,
  #  and add table metadata to a dataframe compiling that info
  # Use suppressWarnings() to avoid "NAs introduced by coercion" message
  EIAdf = readMERtable(sectionW, tableW, onlineReading=T)
  newTableName = paste("MER",sectionW, tableW,sep="_")
  assign(newTableName, EIAdf)

  # # to check if dataframe exists, given name as text string df_name:
  # #  exists(df_name) && is.data.frame(get(df_name))
  # if (!exists("EIAdf_metadata")) {
  #     EIAdf_metadata = data.frame()  
  # } # else append new data to old

  # EIAdf_metadata = rbind(EIAdf_metadata, getMetaData(EIAdf, newTableName ))
  # # returns thge metadata only
}

```

### Get US Oil Production, Consumption, and Import Data

```{r preprocessDataT03_01, echo=echoWorking}
onlineReading = T

sectionWanted = "T03"
tableWanted = "01"
```

```{r}
# following is kind of dumb, because function renames table with EIA MER number,
# for sectionWanted & tableWanted, and that result is renamed when returned.
# SO `readMERtable_and_renam` may not offer any advantage over readMERtable
USOil = readMERtable_and_renam(sectionWanted, tableWanted)
```


```{r readDataT03_01, echo=echoWorking}
# Use suppressWarnings() to avoid "NAs introduced by coercion" message
EIAdf = readMERtable(sectionWanted, tableWanted, onlineReading)
newTableName = paste("MER",sectionWanted, tableWanted,sep="_")
assign(newTableName, EIAdf)

# to check if dataframe exists, given name as text string df_name:
#  exists(df_name) && is.data.frame(get(df_name))
if (!exists("EIAdf_metadata")) {
  EIAdf_metadata = data.frame()  
}
EIAdf_metadata = rbind(EIAdf_metadata, getMetaData(EIAdf, newTableName ))
```


```{r preprocessDataMultTables, echo=echoWorking}
sectionWanted = "T11"
tableWanted = "02"
# Use suppressWarnings() to avoid "NAs introduced by coercion" message
EIAdf = readMERtable(sectionWanted, tableWanted, onlineReading)
newTableName = paste("MER",sectionWanted, tableWanted,sep="_")
assign(newTableName, EIAdf)
# to check if dataframe exists, given name as text string df_name:
#  exists(df_name) && is.data.frame(get(df_name))
if (!exists("EIAdf_metadata")) {
  EIAdf_metadata = data.frame()  
}
if (nrow(EIAdf) > 0)
  EIAdf_metadata = rbind(EIAdf_metadata, getMetaData(EIAdf, newTableName ))

sectionWanted = "T11"
tableWanted = "03"
# Use suppressWarnings() to avoid "NAs introduced by coercion" message
EIAdf = readMERtable(sectionWanted, tableWanted, onlineReading)
newTableName = paste("MER",sectionWanted, tableWanted,sep="_")
assign(newTableName, EIAdf)
# to check if dataframe exists, given name as text string df_name:
#  exists(df_name) && is.data.frame(get(df_name))
if (!exists("EIAdf_metadata")) {
  EIAdf_metadata = data.frame()  
}
if (nrow(EIAdf) > 0)
  EIAdf_metadata = rbind(EIAdf_metadata, getMetaData(EIAdf, newTableName ))


sectionWanted = "T11"
tableWanted = "01B"
# Use suppressWarnings() to avoid "NAs introduced by coercion" message
EIAdf = readMERtable(sectionWanted, tableWanted, onlineReading)
newTableName = paste("MER",sectionWanted, tableWanted,sep="_")
assign(newTableName, EIAdf)
# to check if dataframe exists, given name as text string df_name:
#  exists(df_name) && is.data.frame(get(df_name))
if (!exists("EIAdf_metadata")) {
  EIAdf_metadata = data.frame()  
}
if (nrow(EIAdf) > 0)
  EIAdf_metadata = rbind(EIAdf_metadata, getMetaData(EIAdf, newTableName ))

sectionWanted = "T11"
tableWanted = "01A"
# Use suppressWarnings() to avoid "NAs introduced by coercion" message
EIAdf = readMERtable(sectionWanted, tableWanted, onlineReading)
newTableName = paste("MER",sectionWanted, tableWanted,sep="_")
assign(newTableName, EIAdf)
# to check if dataframe exists, given name as text string df_name:
#  exists(df_name) && is.data.frame(get(df_name))
if (!exists("EIAdf_metadata")) {
  EIAdf_metadata = data.frame()  
}
if (nrow(EIAdf) > 0)
  EIAdf_metadata = rbind(EIAdf_metadata, getMetaData(EIAdf, newTableName ))

# Renewable Energy Production
sectionWanted = "T10"
tableWanted = "01"
# Use suppressWarnings() to avoid "NAs introduced by coercion" message
EIAdf = readMERtable(sectionWanted, tableWanted, onlineReading)
newTableName = paste("MER",sectionWanted, tableWanted,sep="_")
assign(newTableName, EIAdf)
# to check if dataframe exists, given name as text string df_name:
#  exists(df_name) && is.data.frame(get(df_name))
if (!exists("EIAdf_metadata")) {
  EIAdf_metadata = data.frame()  
}
EIAdf_metadata = rbind(EIAdf_metadata, getMetaData(EIAdf, newTableName ))

```

## Get Biofuel Production/Consumption Data and Conversion Factors

```{r combineBiofuelDataTables, echo=echoWorking}
MER_BioFuels = MER_T10_01 %>% filter(MSN=="BFPRBUS")
# https://en.wikipedia.org/wiki/Gasoline_gallon_equivalent
BTUperGGE = 114000 
TrilBTUperYperkBD = BTUperGGE*42*1000*365.25/1E12
kBDperTrilBTUperY = 1/TrilBTUperYperkBD
```


### Combine the Petroleum Data Tables

```{r combinePetroleumTables, echo=echoWorking}
MER_PetData = rbind(MER_T03_01, MER_T11_01A, MER_T11_01B, MER_T11_02, MER_BioFuels)
MER_PetData$YYYYMMc = as.character(MER_PetData$YYYYMM)

MER_PetData = MER_PetData %>% 
  select(-Column_Order) %>% 
  separate(YYYYMMc, sep= 4, into = c("YYYY", "MM")) # split into year and month fields

```

#### look for duplicate variables and drop

```{r dropDuplicateVars, echo=echoWorking}
#  (necessary before spread(), to avoid error)

# Remove duplicate rows based on one or more col values: dplyr::distinct(my_data, colname).
#   distinct() is best-suited for interactive use. 
# The function distinct_() should be used for calling from a function.
# In this case the input must be “quoted”. 
# Also can use base fns unique() and duplicated()
# MER_PetData = MER_PetData %>%
#  distinct_("MSN", "Value", "YYYYMM")  # leaves only the variables tested for distinctness

Pet = MER_PetData %>% select(MSN, Value, YYYYMM)

MER_Dupl_PetData = MER_PetData[duplicated(Pet), ]  # keep all vars, if distinct name/value/date
MER_PetData = MER_PetData[!duplicated(Pet), ]  # keep all vars, if distinct name/value/date
```

get annual data subset

```{r extractAnnualSubset, echo=echoWorking}
MER_PetDataA = MER_PetData %>% filter(MM == "13") # get annual data only

# MER_PetDataA = MER_PetDataA %>%
#  distinct_("MSN", "Value", "YYYYMM")

```

Keep only essential petroluem production/consumption fields (name MSN, Value, and date )
 and spread out vars into columns.

Note that petroleum consumption equals "Petroleum Products Supplied."


```{r spreadMainPetrVars, echo=echoWorking}

varsWanted = c("PAPRPUS", "NLPRPUS", "PNPRPUS",         # Field prod: Total Crude, NGL, Total
               "RYPRPUS",                               # Renewable Fuels and Oxygenates
               "PAGLPUS",                               # Petr Processing Gain
               "PAEXPUS", "PAIMPUS", "PANIPUS",         # Exports, Imports, Net Imports
               "PASCPUS",                               # Petr Stock Change
               "PAAJPUS",                               # Adjustments
               "PATCPUS",                               # Petr Products Supplies
               "BFPRBUS"                                # Biofuels Production
)
EIAdf_metadata %>% filter(MSN %in% varsWanted)
# PATCPUS = Petroleum Consumption, United States = Petroleum Products Supplied
# PAPRPUS = Crude Oil Production, United States = Crude Oil Production, Total

# Petroleum Consumption Series From Tables 3.7A, B, C
# c("PARCPUS", "PACCPUS", "PAICPUS","PAACPUS", "PAEIZUS")
# Residential, Commercial, Industrial, Transportation and Electric


PetA = MER_PetDataA %>% 
  select(-c(YYYYMM, Description, Unit, MM)) %>%
  filter(MSN %in% varsWanted) %>%
  spread(key = MSN, value=Value)

PetA$BFPRBUS = PetA$BFPRBUS*kBDperTrilBTUperY
```

```{r reportSelectedPetrValues, echo=echoWorking}
rownames(PetA) = PetA$YYYY
cat("2007 Consumption: ", PetA["2007", "PATCPUS"], "\n")
cat("2012 Consumption: ", PetA["2012", "PATCPUS"], "\n")
cat("2017 Consumption: ", PetA["2017", "PATCPUS"], "\n")

cat("2007 Production: ", PetA["2007", "PATCPUS"], "\n")
cat("2012 Production: ", PetA["2012", "PATCPUS"], "\n")
cat("2016 Production: ", PetA["2016", "PATCPUS"], "\n")
cat("2017 Production: ", PetA["2017", "PATCPUS"], "\n")

cat("2007 Biofuel Prod: ", PetA["2007", "BFPRBUS"], "\n")
cat("2017 Biofuel Prod: ", PetA["2017", "BFPRBUS"], "\n")

# test whether Petroleum Net Imports (PANIPUS) matches Imports - Exports
sum(abs(PetA$PAIMPUS - PetA$PAEXPUS - PetA$PANIPUS))
test = PetA %>% 
  mutate(
    NetImpBal = PAIMPUS - PAEXPUS - PANIPUS,
    FldPrdBal = PAPRPUS + NLPRPUS - PNPRPUS,
    SupTotal  = PNPRPUS + PAGLPUS + RYPRPUS + PANIPUS - PASCPUS + PAAJPUS,
    SupDemBal = PNPRPUS + PAGLPUS + RYPRPUS + PANIPUS - PASCPUS + PAAJPUS - PATCPUS
  )

PetA = PetA %>% 
  mutate(
    ImpFract  = PANIPUS/PATCPUS,
    YYYY = as.numeric(YYYY)
  ) %>%
  rename(Year = YYYY)

rownames(PetA) = PetA$Year
cat("2007 Imports/Consumption: ", PetA["2007", "ImpFract"], "\n")
cat("2016 Imports/Consumption: ", PetA["2016", "ImpFract"], "\n")
cat("2017 Imports/Consumption: ", PetA["2017", "ImpFract"], "\n")
cat("2018 Imports/Consumption: ", PetA["2018", "ImpFract"], "\n")
```

Visualize the data
-------------------

#### U.S. Petroleum Production
```{r visualizeAnnualProdData0, echo=echoWorking}
PetA %>% select(-ImpFract) %>%
  gather(key=MSN, value=Value, -Year) %>%
  ggplot(aes(x=Year, y=Value, color=MSN))+geom_line() +
  ggtitle("US Petroleum Supplies")
```

```{r tableOfSupplySeriesMetadata, echo=echoWorking}
EIAdf_metadata %>% filter(MSN %in% varsWanted) %>%
  select(MSN, Description) %>%
  arrange(as.character(MSN)) %>%
  kable(caption = "Table: Legend for Petroleum and Other Liquids Supply")
```

#### U.S. Net Imports as a Fraction of Consumption
```{r visualizeAnnualData0, echo=echoWorking}
PetA %>%
  ungroup() %>%
  ggplot(aes(x=Year, y=ImpFract))+geom_line() +
  ggtitle("US Petroleum Net Imports as a Fraction of Consumption")

# PetA %>% select(Year,ImpFract) %>% str()
```


### Plot Monthly Petroleum Consumption

``` {r plotPetrConsumption, echo=echoWorking}
PetrCons = MER_T11_02
PetrCons$Date <- as.Date(paste0(as.character(PetrCons$YYYYMM),"15"),format="%Y%m%d") # returns NA for month 13 (annual obs)
PetrCons = PetrCons %>% separate(Description, sep=",", c("Series", "Region"))

PetrCons %>%  filter(!is.na(Date)) %>% # drops annual obs
  filter(Series=="Petroleum Consumption" & !is.na(Value)) %>%
  ggplot(aes(x=Date, y=Value, color=Region)) + 
  geom_line() + ggtitle("Petroleum Consumption (Monthly Data)") + ylab("1000 B/day")

```

```{r showPetrConsSpread, echo=echoWorking}
# PetrCons %>%  filter(!is.na(Date)) %>% # drops annual obs
#   filter(Series=="Petroleum Consumption", !is.na(Value)) %>%
#   select(Date, Value, Region) %>%
#   spread(key=Region, value = Value)
```

### Prices

    9.1 Crude Oil Price Summary
    9.2 F.O.B. Costs of Crude Oil Imports From Selected Countries
    9.3 Landed Costs of Crude Oil Imports From Selected Countries
    9.4 Retail Motor Gasoline and On-Highway Diesel Fuel Prices
    9.6 Refiner Prices of Petroleum Products for Resale
    9.7 Refiner Prices of Petroleum Products to End Users
    9.10 Natural Gas Prices
    10.1 Renewable Energy Production and Consumption by Source
    10.2 Renewable Energy Consumption
    10.3 Fuel Ethanol Overview

```{r readMERPrices, echo=echoWorking}

# Petroleum Prices - Crude Oil Price Summary
sectionWanted = "T09" # Petroleum Prices
tableWanted = "01"    # Crude Oil Price Summary
# Use suppressWarnings() to avoid "NAs introduced by coercion" message
EIAdf = readMERtable(sectionWanted, tableWanted, onlineReading)
newTableName = paste("MER",sectionWanted, tableWanted,sep="_")
assign(newTableName, EIAdf)
# to check if dataframe exists, given name as text string df_name:
#  exists(df_name) && is.data.frame(get(df_name))
if (!exists("EIAdf_metadata")) {
  EIAdf_metadata = data.frame()  
}
EIAdf_metadata = rbind(EIAdf_metadata, getMetaData(EIAdf, newTableName ))
```

#### Note: Reading GDP and Deflators does not yet work right due to different convention for naming tables in Appendices.

Seek Appendix Table TC1 : Population, US GDP and US Gross Output

```{r readMERGDPandDeflators, echo=echoWorking}
# Table C1. Population, U.S. Gross Domestic Product, and U.S. Gross Output
sectionWanted = "TC1" # Appendix (section and table not separated by ".")
tableWanted = ""      # signal that section and table not separated by "."
# https://www.eia.gov/totalenergy/data/browser/csv.php?tbl=TC1
# vs
# "http://www.eia.gov/totalenergy/data/browser/csv.cfm?tbl=T01.03" # MER Table 1.3 Primary Energy Consumption by Source

# Use suppressWarnings() to avoid "NAs introduced by coercion" message
EIAdf = readMERtable(sectionWanted, tableWanted, onlineReading)
newTableName = paste("MER",sectionWanted, sep="_")
assign(newTableName, EIAdf)
# to check if dataframe exists, given name as text string df_name:
#  exists(df_name) && is.data.frame(get(df_name))
if (!exists("EIAdf_metadata")) {
  EIAdf_metadata = data.frame()  
}
EIAdf_metadata = rbind(EIAdf_metadata, getMetaData(EIAdf, newTableName ))
```

### Plot Monthly OECD Stocks

``` {r plotPetrStocks, echo=echoWorking}
PetrStocks = MER_T11_03
PetrStocks$Date <- as.Date(paste0(as.character(PetrStocks$YYYYMM),"15"),format="%Y%m%d")
PetrStocks = PetrStocks %>% separate(Description, sep=",", c("Series", "Region"))

PetrStocks %>% filter(!is.na(Date)) %>% # drops annual obs
  filter(Series=="Petroleum Stocks", !is.na(Value)) %>%
  ggplot(aes(x=Date, y=Value, color=Region)) + 
  geom_line() + ggtitle("Petroleum Stocks") + ylab("Million Barrels")

```

### Catalog of Series Loaded

```{r catalogSeriesMetadata, echo=echoWorking}
EIAdf_metadata %>% select(-Column_Order) %>%
  kable(caption = "Metadata for Primary Series Downloaded")
```

