---
title: "test EIA MER byAPI"
author: "Paul Leiby"
date: "3/2/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

- [eia: API Wrapper for 'US Energy Information Administration' Open Data](https://cran.r-project.org/web/packages/eia/index.html)
- [API Details](https://cran.r-project.org/web/packages/eia/vignettes/api.html)
- [EIA categories](https://cran.r-project.org/web/packages/eia/vignettes/categories.html)

It is important to store your API key somewhere secure. Do not commit it to a repository or otherwise share it. For example, you could store it in your .Renviron file.

```{r API_key}
# Your API key for 
username <- "leibypn@outlook.com"
# is
my_API_key <- "BdZfwDN9ZWzkYvjFqexq1HqzjTacAcZBXmK3fBmg"
```

Note: As of February 11th, 2022: The EIA API now only respond to HTTPS. Insecure (HTTP) web protocols are no longer supported. 


```{r}
library(tidyverse)
library(eia)
```

```{r}
# If you need to set it, you can do so as follows.
eia_set_key(my_API_key) # store your key for the duration of the R session.
# eia_get_key() # retrieve it
```


```{r}
key <- Sys.getenv("EIA_KEY") 
# or:
key <- eia_get_key() # for key in local environment
```

```{r}

system.time(eia_cat <- eia_cats()) # top-level category information. API call; cache result
system.time(eia_cats()) # will read from cache
system.time(eia_cats(cache = FALSE)) # force API call
```

```{r}
eia_clear_cats() # clear the cache for the catalog
```

```{r}
system.time(x <- eia_parent_cats(742, cache = FALSE)) # View all parent categories for a given child category.

```

```{r}
x
```

```{r}
eia_child_cats(389) # View the immediate child categories for a given parent category.
# **Output format:** The default is to return tidy data in a tibble data frame.
# For eia_cats you can set 
#   tidy = FALSE to return the list returned by jsonlite::fromJSON without any further processing
#   or tidy = NA to return the raw JSON data as a character string.

eia_cats(tidy = FALSE)
```

### EIA time series data

Time series data is obtained by series ID. Most columns contain metadata. The data column contains the time series data.

```{r}
id <- "ELEC.GEN.ALL-AK-99.A"
(x <- eia_series(id, start = 2010, end = 2019))
```

```{r}
# The `data` column includes the time series data
x
```

```{r}
x_df = x$data
x_df
```

```{r}
select(x, units, data) %>% unnest(cols = data) %>%
  ggplot(aes(factor(year), value)) + geom_col() +
  labs(x = "Year", y = x$units[1], title = "Net Alaska electricity generation",
       subtitle = "From all fuels", caption = x$description[1])
```

You can provide arguments like the following:

    eia_series(id) # max results
    eia_series(id, n = 5) # most recent five
    eia_series(id, end = 2016, n = 5) # ending in 2016
    eia_series(id, start = 2000, end = 2016) # specific period

As with eia_cats, the output format does not need to be tidy:

    eia_series(id, n = 5, tidy = FALSE) # results of jsonlite::fromJSON
    eia_series(id, n = 5, tidy = NA) # origina JSON as character string

This allows you to use the returned results with existing code you may have that requires data in one of these less processed structures.

### EIA geosets

Geosets are metadata structures organizing time series datasets that can be mapped. Arguments to `eia_geoset` are the same as eia_series with the addition of `region`. Like `id`, `region` can be a vector. Most of the details are the same as before.

In the example below using total electricity generation, get the last two data points for each of and two US states. dplyr and tidyr are used here to clean up the result a bit for purposes of display. gpplot2 is used to graph the data after it has been unnested for each state.

```{r}

id <- c("ELEC.GEN.ALL-99.M") # monthly
region <- c("USA-CA", "USA-NY")
(x <- eia_geoset(id, region, start = "201801", end = "201812"))
#> # A tibble: 2 x 11
#>   geoset_id       setname                             f     units            series_id        name                                       region latlon start end   data        
#>   <chr>           <chr>                               <chr> <chr>            <chr>            <chr>                                      <chr>  <chr>  <chr> <chr> <list>      
#> 1 ELEC.GEN.ALL-9~ Net generation : all fuels : all s~ M     thousand megawa~ ELEC.GEN.ALL-CA~ Net generation : all fuels : California :~ USA-CA <NA>   2001~ 2020~ <tibble [12~
#> 2 ELEC.GEN.ALL-9~ Net generation : all fuels : all s~ M     thousand megawa~ ELEC.GEN.ALL-NY~ Net generation : all fuels : New York : a~ USA-NY <NA>   2001~ 2020~ <tibble [12~

select(x, region, data) %>% unnest(cols = data)
#> # A tibble: 24 x 5
#>    region  value date        year month
#>    <chr>   <dbl> <date>     <int> <int>
#>  1 USA-CA 14364. 2018-12-01  2018    12
#>  2 USA-CA 14605. 2018-11-01  2018    11
#>  3 USA-CA 16864. 2018-10-01  2018    10
#>  4 USA-CA 17592. 2018-09-01  2018     9
#>  5 USA-CA 20994. 2018-08-01  2018     8
#>  6 USA-CA 22356. 2018-07-01  2018     7
#>  7 USA-CA 17514. 2018-06-01  2018     6
#>  8 USA-CA 15932. 2018-05-01  2018     5
#>  9 USA-CA 14811. 2018-04-01  2018     4
#> 10 USA-CA 14039. 2018-03-01  2018     3
#> # ... with 14 more rows

unnest(x, cols = data) %>%
  ggplot(aes(date, value, color = region)) + geom_line() +
  labs(y = x$units[1], title = "Net electricity generation", subtitle = "From all fuels")

## plot of chunk geoset1

```


Another convenience of eia_geoset is the ability to provide regions in the following forms.

    2-character US state abbreviations
    State names
    US Census region names
    US census division names

These shortcuts make it easier to construct an API call involving several states.

```{r}
region <- c("AK", "New England")
x <- eia_geoset(id, region, n = 2)
select(x, region, data) %>% unnest(cols = data)
#> # A tibble: 14 x 5
#>    region value date        year month
#>    <chr>  <dbl> <date>     <int> <int>
#>  1 USA-AK  534. 2020-08-01  2020     8
#>  2 USA-AK  626. 2020-07-01  2020     7
#>  3 USA-CT 4001. 2020-08-01  2020     8
#>  4 USA-CT 4443. 2020-07-01  2020     7
#>  5 USA-MA 2200. 2020-08-01  2020     8
#>  6 USA-MA 2754. 2020-07-01  2020     7
#>  7 USA-ME  861. 2020-08-01  2020     8
#>  8 USA-ME  962. 2020-07-01  2020     7
#>  9 USA-NH 1632. 2020-08-01  2020     8
#> 10 USA-NH 1794. 2020-07-01  2020     7
#> 11 USA-RI  889. 2020-08-01  2020     8
#> 12 USA-RI  845. 2020-07-01  2020     7
#> 13 USA-VT  194. 2020-08-01  2020     8
#> 14 USA-VT  203. 2020-07-01  2020     7

region <- "Middle Atlantic"
x <- eia_geoset(id, region, n = 12)
select(x, region, data) %>% unnest(cols = data)
#> # A tibble: 36 x 5
#>    region value date        year month
#>    <chr>  <dbl> <date>     <int> <int>
#>  1 USA-NJ 6639. 2020-08-01  2020     8
#>  2 USA-NJ 7385. 2020-07-01  2020     7
#>  3 USA-NJ 5721. 2020-06-01  2020     6
#>  4 USA-NJ 4169. 2020-05-01  2020     5
#>  5 USA-NJ 4334. 2020-04-01  2020     4
#>  6 USA-NJ 4469. 2020-03-01  2020     3
#>  7 USA-NJ 4912. 2020-02-01  2020     2
#>  8 USA-NJ 5488. 2020-01-01  2020     1
#>  9 USA-NJ 6193. 2019-12-01  2019    12
#> 10 USA-NJ 5510. 2019-11-01  2019    11
#> # ... with 26 more rows

unnest(x, cols = data) %>%
  ggplot(aes(date, value, color = region)) + geom_line() +
  labs(y = x$units[1], title = "Net electricity generation", subtitle = "From all fuels")
```